---
title: "On the Use of R for Machine Learning"
author: "Avik Mohan"
output: html_document
---

R employs a number of different data structures to contain the data to be analyzed. Some examples of data structures are arrays, sets, and trees.

The data sets we will use in R are mainly vectors, factors, lists, data frames, and matrices.
For this section, we will be using the "01a_data.txt" file from the biomed data.
I keep these folders locally in my ~/AAUCLA/SOCR/data folder, but rename the path of the file to fit your directory.

You can read this file into R after downloading it by using:
```{r}
`patientData` <- read.delim("~/AAUCLA/SOCR/data/01a_data.txt")
```

The read.delim() function reads in the dataset to the specified variable, deliminating the data using spaces by default.

We can explore this data to learn about these data structures.

We can access the ith index of our data by using data[i].

Also one can select a range for one of the indexes, ranging from 1 to 10, using the notation 1:10.

Finally, leaving an index blank tells R that you want to include everything for that index. 

Putting this together, we can try:

```{r}
patientData[1:10,]
```
To obatin all columns for the first 10 people.

Now we take

```{r}
patientData[1:10,]
patientData[0,]
patientData[1:10,3]
patientData[1:10,4]
```

We can see here that we have a number of traits recorded about baseball players. These traits are represented by different data types. Position is a factor, and weight is a numeric.

These types are held in vectors that we pull from the raw data. Lets look at what a vector is.

##Vectors
	
Vectors act as the R analog to an array. Vectors are ordered sets of elements. For example, we can store the types integer, numeric (decimal), alphabetical character, or bools into a vector. Note that we cannot mix types in these vectors (we will see how to do this with lists).

We can create our first vector using the following notation.
```{r}
temperature <- c(98.1, 98.6, 101.4)
temperature[3]
```

Notice that vectors are 1-indexed, not 0-indexed. That is, the first column of information, Name, is accessed for the first person using
```{r}
patientData[1,1]
```
Rather than
```{r}
patientData[0,0]
```

Now lets look at a subset of the of the Name data

```{r}
names <- patientData[1:15, 1]
names
```

We can exclude certain elements:

```{r}
names[-2]
names[-(2:3)]
names[-(c(2,3))]
```

And can also use boolean values to access elements.
Here, we specifiy not to include the third element.
```{r}
names[c(TRUE, TRUE, FALSE)]
```

##Factors

We now show how to represent 'nominal' or categorical data. We could use a vector of characters, we have factors specifically for this. Factors are a special subtype of vectors. We use factors since they are more efficient (they store each category type only once) and they work nicer with some algorithms that use special routines to handle categorical variables.

Lets create our first factor:
```{r}
gender <- factor(c("MALE", "FEMALE", "MALE"))
gender
gender[2]
```

When R displays the 'levels', it is displaying the full set of  categories the data contains.

If our data doesn't contain all possible levels, we can add them ourselves:

```{r}
grades <- factor(c("A", "B", "A", "C", "A"), 
                 levels = c("A", "B", "C", "D", "F"))
grades
```

Notice that even though every grade type wasn't actually given (no one got a D or an F) we still recognize these as valid possible values for grade.

##Lists

Lists are another special type of vector. Unlike vectors however, we can store mixed types in lists. 

Consider our patientData. Suppose we wanted all the information on the first person:
```{r}
patientData[1,]
```

We can see that there are a number of different data types displayed. Suppose we wanted to be able to refer to this person as their own separate object. We could make a list to hold this information:

```{r}
subject1 <- list(fullname = patientData[1,1], team = patientData[1,2], position = patientData[1,3],
                height = patientData[1,4], weight = patientData[1,5], age = patientData[1,6])
subject1
```

The command is often useful to group data together when it may come from different sources. 

We can access lists as usual:
```{r}
subject1[1]
```

We could use these lists to represent our entire dataset, but this is a common enough endeavour that R provides a dedicated structure.

##Data Frames


Data frames are analogous to spreadsheets or databases. It contains both rows and columns. We can interpret these data frames as lists of vectors or factor, each with the same number of values.

To simulate creating a data frame from raw data, lets strip apart our patientData into its composing parts, and pretend we were given these to start:

```{r}
names <- patientData[,1]
teams <- patientData[,2]
positions <- patientData[,3]
heights <- patientData[,4]
weights <- patientData[,5]
ages <- patientData[,6]

ourDataFrame <- data.frame(names, teams, positions, heights, weights, ages)
ourDataFrame[1:15,]
```

We can see that we end up with something very similar to the imported patientData. You can use data frames to create your own datasets from raw data when not using the ones from the SOCR database.

We can also access data frames in interesting ways:
```{r}
ourDataFrame$ages[1:15]
ourDataFrame[1:10, c("weights", "ages")]
```

You can also use a negative index to exclude an element, as we had for vectors.

##Matrices

Matrices are similar to data frames in that they hold tabular data. Matrices can hold any single type of data, though the type is usually numeric as matrices are often used for mathematical purposes.

Creating a matrix requires a bit of understanding of how the indexing works.

Lets try to make a matrix to see what I mean:
```{r}
m <- matrix(c(17, 32, 33, 21))
m
m <- matrix(c(17, 32, 33, 21), nrow = 1)
m
m <- matrix(c(17, 32, 33, 21), nrow = 2)
m
m <- matrix(c(17, 32, 33, 21), ncol = 2)
m
```

We can see here that the matrices default to a column vector if rows or columns is not specified. Additionally, the matrix fills out the first column of the matrix first, then proceeds to fill in the second column, and so on. This is called column-major order. We can look at this more closely:


```{r}
m <- matrix(c(17, 32, 33, 21, 13, 19), nrow = 2)
m
```

A row-major order would have filled out the matrix as:

17 \     32   \   33 \ 


21  \    13   \   19

In addition to matrices are arrays, which are matrices which can have multiple layers, e.g. higher dimension. For the techniques examined here, this data type is beyond our scope.

A review of the data structures here and a brief exploration of other R data structures can be found at
http://adv-r.had.co.nz/Data-structures.html
and also at
https://ramnathv.github.io/pycon2014-r/learn/structures.html